Can historical data predict population responses to climate change experiments?

### 2016-12-18 precip/manuscript/precip_ms.tex
adding results text, need to add figures and tables next

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016, adler_in_prep}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\label{eqn:wijm}
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

@@

We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}. Each model was run for 2,000 iterations and four independent chains with different initial values for parameters. We discarded the initial 1,000 samples. Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.

@@

\subsection*{Individual-based models}

The vital rate regressions allow us to evaluate whether the experimental treatments and the soil moisture variables considered had an effect on species performance.  But the population response of each species ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.  To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next in the experimental plots.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.

For control plots, we used a version of the model with all removal treatment effects set to zero. For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero.  We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.

\subsection*{Assessing soil moisture effects and predictive accuracy}

After fitting the random year effects (null) and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots. We then assessed performance of the climate and null models by calculating the mean square error between the predictions and observations for each of the vital rates. In addition we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.  We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.  The average of this product across all the years of the experiment then gives the predicted treatment coefficient for treatment level $h$. These predicted treatment effects can then be compared directly to the treatment effects fitted in the treatment model in the term $\chi$.

For treatment and vital rate models we considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.

We evaluated cover predictions from the climate and year effects model similarly.  We first fit the

@@

During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a). However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (\ref{fig:climate}b).

The experimental drought shelter successfully reduced the soil moisture in the study plots in the spring and summer, while irrigation successfully increased soil moisture \ref{fig:spotm}. Our continuous data

\subsection*{Cover trends}

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

Our experiment showed that including climate covariates in our population models did little to improve population level predictions across the course of the experiment.  In general the climate models did not perform as well as null models in predicting response to the experimental drought and irrigation treatments at the vital rate or the population level.  However, there were some glimmers of hope.  First at the population level we generally found that including climate covariates did not improve predictions in the control and irrigation plots, but they did appear to improve predictions in the drought plots ().  This suggests that models fit to observational data do contain useful information about the effects of climate on individual performance.  The fact that these predictions showed up in the more extreme conditions produced by the drought plots could be a sign that in ambient conditions the effect of climate is weak enough that it is washed out at the population level by finer scale sources of variation. This pattern was observed for species x, x and x.

>>>>

Although the main focus of the current analysis the effects of soil moisture, we also modeled the effects of inter- and intra-specific competition in our vital rate models.  We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016, adler_in_prep}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\label{eqn:wijm}
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

@@

We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}. The prior distributions and model code used are described more completely in appendix A. Each model was run for 2,000 iterations and four independent chains with different initial values for parameters. We discarded the initial 1,000 samples. Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.

@@

\subsection*{Predicting cover from individual-based models}

The vital rate regressions allow us to evaluate whether soil moisture and the experimental treatments had an effect on species performance.  But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.  To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size, defined as plot-level percent cover, from one year to the next in the experimental plots.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the observed genet sizes and locations of the four focal species observed in year $t$ in each quadrat. For every individual genet in a quadrat, we projected its size and survival probability in the next year using the growth and survival models and the appropriate crowding and soil moisture or treatment covariates for that year and quadrat.  Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model. We calculated the expected cover in year $t+1$ as the sum area of new recruits, plus the sum of the expected area of existing plants multiplied by each plant's expected survival probability.

We accounted for the uncertainty in our random year effects in our projections, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}. We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).

We generated predictions for the full time series of observations, including the five experimental years, from the year effects model, the climate model and the treatment model. After generating predictions for each year, we found the mean cover across all quadrats in each treatment level and then calculated the predicted log cover change as $log(Cover_{t+1}/Cover_{t})$.

\subsection*{Quantifying predictive accuracy}

After fitting the year effects and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots. We then assessed performance of the climate and null models by calculating the mean square error (MSE) between the predicted and observed responses in the experimental data as,

\begin{equation}
MSE = \frac{1}{n} \sum_{i=1}^{n} (y_i - E(y_i|\theta))^2, 
\label{eqn:MSE}
\end{equation}
where $y_i$ is the outcome of observation $i$ and $E(y_i|\theta)$ gives the expected outcome given the parameters in the model $\theta$. The MSE is easy to interpret, but is not always appropriate for models fit with non-normal error structures \citep{gelman_understanding_2014}. A more general statistic for assessing model predictions is the log pointwise predictive density (lppd) \citep{gelman_understanding_2014}.  The lppd for a given model is defined as,

\begin{equation}
lppd = \sum_{i=1}^{n} log \int p(y_i| \theta)p_{post}(\theta) d\theta, 
\label{eqn:lppd}
\end{equation}
where the integral on the right side gives the probability of observing the outcome $y$ at each data point $i$ given the full posterior distribution of the parameters in the model $p_{post}(\theta)$. In practice we computed the lppd from the posterior simulations generated by STAN as,

\begin{equation}
\widehat{lppd} = \sum_{i=1}^{n} log \left(\frac{1}{S} \sum_{s=1}^{S} p(y_i | \theta^S) \right),
\label{eqn:clppd}
\end{equation}
where the summation of $p(y_i|\theta^S)$ gives the total probability of observing the the actual response $y_i$ given the simulated posterior distribution $\theta^S$ across the full set of model simulations $S$.  The log of this sum is then averaged across the set of all observations $i$.  Higher lppd scores indicate that the model better predicts the observations.

In addition, we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.  We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.  We then averaged these across all five years in the experiment to find the average treatment effect predicted by the climate model.  We compared these to the posteriors of the treatment parameters, $chi$, from the treatment model.  As a measure of agreement between our predictions and observed response we calculated the correlation between the predicted and observed parameter values.

We considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.

@@

\subsection*{Effects on soil moisture}

We successfully changed the average soil moisture measured in the plots we covered with drought shelters and irrigated, in the directions we expected (fig); our spot measurements conducted during the springs of X, X and X showed that drought plots were X percent drier, while irrigation plots were X percent wetter than ambient plots.  This difference was highly significant () but there was a lot of spatial variation in the soil moisture measurements within treatments (\ref{fig:}). The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently (). The best model for predicting the average treatment effects on soil moisture was : .  We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().

During the course of our experiment annual precipitation recorded at the local climate station varied from X to X percent of period of record averages (1925 to 2016).  The year of X was exceptionally dry while the year of X was.  The precipitation during the summer of X was the wettest on record.  Mean annual average air temperature during this period were consistently well above average for the period of record, with the years of X X and X being among the warmest ever recorded at this site.

The SOILWAT soil moisture model indicated that soil moisture fluctated from between X and X on average each year, with the months of X and X being the wettest while soil moisture during the months of X and X was driest (fig). Predicted seasonal soil moisture for each year was positively correlated with precipitation, and negatively correlated with average temperatures. The average soil moisture predicted from soilwat for the entire study period from X to X is shown in figure X.  Soil moisture predictions for the ambient conditions from 2011 to 2016 fall on the low side of the period of record averages.  This was driven by higher than average temperatures and severely dry winters and spring conditions in the years of X.

When we adjust the seasonally averaged soil moisture predictions by the treatment effects, we find that the soil moisture conditions predicted in the drought plots during the course of the experiment are lower than those predicted for any of the previous years in the historical record.  Likewise the predicted soil moisture in the irrigation plots was far above the average for the historical record in the years of X and X.  We interpret this as evidence that our drought and irrigation plots created truly novel soil moisture conditions in the years X and X that were unlikely to have ever been experienced by the study species during the entire period of the observational data. We visualize this by conducting a principle component analysis (PCA) on all the seasonal soil moisture, and plotting each year on the first two principle components. The PCA shows that the ambient, drought and irrigation plots experienced warmer than average seasonal conditions during the course of the experiment (further left on PCA1).  The drought plots were considerably drier, and the irrigated plots considerably wetter (figure PCA).

\subsection*{Effects on cover and vital rates}

The cover of \textit{H. comata} and \textit{P. spicata} fell significantly by X and X percent respectively from the pre-treatment cover in 2011 to 2016. The cover of \textit{P. secunda} showed a trend towards decrease in the drought plots and an increase in the irrigated plots but these changes were not significant.  In contrast the cover of \textit{A. tripartita} changed in the opposite direction from the grasses; it increased in the drought plots and decreased in the irrigated plots but these changes were not significant (fig).

Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species (table x). Among the survival models, WAIC scores indicated that we should retain size x treatment effects for \textit{P. secunda}.  Among the growth models, we retained size by treatment effects for \textit{A. tripartita} and \textit{P. secunda}. DISCUSS SIGNIFICANT treatment effects: For \textit{A. tripartita} we observed a significant decrease in growth in the x treatment, while we observed no change in the X treatment.  \textit{H. comata} growth, survival and recruitment, while growth. \textit{P. secunda} growth and survival recruitment.  \textit{P. spicata} growth survival and recruitment. Treatment by size effects on \textit{A. tripartita} and \textit{P. secunda} showed that ()

We observed a number of strong correlations between the year effects in the vital rate models and the seasonal soil moisture variables and table x reports the strongest three correlations for each species and vital rate (table). In general the soil moisture during the spring and summer period appeared to be more strongly correlated with the year effects for each species than the fall soil moisture, but see \textit{P. spicata} vr.  All three time lags show up in the choosen variables, but 1 out of X variables choosen involved soil moisture in the year immediately before the first year of the transition, (subscript "0"), while variables involving soil moisture in the year before that (subscript "lag") and after that (subscript "1") were choosen less often.

We observed vital rate and species-specific effects of soil moisture on each of these species after fitting the vital rate models with the soil moisture covariates discussed above (fig). DISSCUSS significant soil moisture effects: for \textit{A. tripartita} growth was significantly affected by ...  For \textit{H. comata}.  For \textit{P. secunda}. And for \textit{P. spicata}. Soil moisture by size effects on \textit{A. tripartita} and \textit{P. secunda} showed that ()

\subsection*{Evaluating the predictions}

For each vital rate and species, predictions from the year effects and climate models were nearly indistinguishable (fig). This indicated that the additional soil moisture covariates in the climate models had very little effect on the predictions. Not surprisingly \citep{chu_direct_2016,treddenick_} our models show that plant size and intra-specific crowding are far more influential in determining growth, survival and recruitment than climate effects.  In all cases, we found MSE and lppd scores for the climate models were slightly worse than the MSE and the lppd of the year effects model (table).  This indicated that adding the soil moisture parameters did not improve model predictions over all (table). When we break the data up into treatment groups we do see that the climate model outperformed the year effects model for predicting X species VR in treatment T.

While our climate models failed to improve predictions, we did find that they predicted treatment responses that were largely in the same direction as the observed treatment responses (fig).  The overall correlation between the predicted treatment parameters and the observed treatment parameters was X with stronger correlation for the climate effects on the size by treatment effects fit for \textit{P. secunda} growth and survival and \textit{A. tripartita} growth (fig).  A table of MSE between predicted and expected treatment parameters shows that averaged over all species and vital rates, we did the best job of predicting the response to the drought treatment.  Averaged across all treatements and vital rates, we did the best job predicting the response of \textit{species} to the treatments. Finally if we average across all species and treatments, we did the best job of predicting VR parameters in response to the treatment.

As for the vital rate predictions, our one-step ahead cover predictions from the year effects models and the climate models were fairly similar (fig). Models with and without soil moisture parameters tended to underpredict the cover of \textit{A. tripartita} and \textit{P.secunda} during the experiment (fig).  When we compare the MSE of the predicted population growth rates for each year and quadrat from the model with and without the soil moisture parameters, we find that adding soil moisture to the vital rate models always made the one step ahead cover predictions slightly less accurate (table).  The pattern when we consider the correlation between the predicted and observed population growth rates was slightly different (table?).  When we broke the population growth rate data up by treatment, we found that adding soil moisture to the models mostly did not improve predictions of cover change in the control and irrigation plots.  However, including the soil moisture parameters led to better predictions of changes in population growth rates for three of the species (S , S, and S) in the drought plots (fig).

As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.

(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.01, 0.65) and
 \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

Our experiment showed that including climate covariates in our population models did little to improve population level predictions across the course of the experiment.  Averaged across the control, drought and irrigation treatments, the climate models did not perform as well as null models in predicting response the experiment at the vital rate or at the level of population growth rates.  However, there were some glimmers of hope.  First at the population level we generally found that including climate covariates did not improve predictions in the control and irrigation plots, but they did appear to improve predictions in the drought plots ().  This suggests that models fit to observational data do contain useful information about the effects of climate on individual performance.  The fact that these predictions showed up in the more extreme conditions produced by the drought plots could be a sign that in ambient conditions the effect of climate is weak enough that it is washed out at the population level by finer scale sources of variation. This pattern was observed for species x, x and x.
